Design and Implementation of Probabilistic Programming Language Anglican

### 2016-07-05 paper/paper.tex
some citations

While probabilistic programming systems shorten the iteration cycle in
exploratory model design, they typically lack basic functionality needed for
data I/O, pre-processing, and analysis and visualization of inference results.
In this demonstration, we describe the implementation of
Anglican~(\url{http://www.robots.ox.ac.uk/~fwood/anglican/}), a probabilistic
programming language that tightly integrates with
Clojure~(\url{http://clojure.org/}), a general-purpose programming language
that runs on the Java Virtual Machine (JVM). Both languages share a common
syntax, and can be invoked from each other. This allows Anglican programs to
make use of a rich set of libraries written in both Clojure and Java.
Conversely Anglican allows intuitive and compact specification of models for
which inference may be performed as part of a larger Clojure project.

Anglican is a probabilistic programming language integrated with
Clojure.  There are several ways to build a programming language
on top or besides another language.  The easiest to grasp is an
interpreter --- a program that reads a program, in its entirety
or line-by-line, and executes it by applying operational
semantics of a certain kind to the language. \textsc{Basic} is
famous for line-by-line interpreted implementations.

@@

An implementation of Anglican must therefore address three issues:
\begin{itemize}
    \item the Clojure syntax to introduce probabilistic Anglican
        code inside Clojure modules;
    \item source-to-source transformation of Anglican programs
        into Clojure, so that probabilistic execution becomes
        possible;
    \item algorithms which run Clojure code, obtained by
        transforming Anglican programs, according to the
        probabilistic operational semantics.
\end{itemize}
The following sections explain the way Anglican is implemented,
from source-to-source transformation and syntactic wrappers to
inference algorithms which accept Clojure functions built from
Anglican code as a parameter, and produce probabilistic outputs.

@@

An Anglican program, or \textit{query}, is compiled into a Clojure
function. When inference is performed with a provided algorithm, this
produces a sequence of return values, or \textit{predicts}. Anglican
shares a common syntax with Clojure; Clojure functions can be called
from Anglican code and vice versa. A simple program in Anglican can
look like the following code:
\begin{lstlisting}[style=default]
(defquery models
  "chooses a distribution
   which describes the data"
  (let [;; Model - randomly choose distribution
	    dist (sample (categorical
                       [[uniform-continuous 1]
                        [normal 1] 
                        [gamma 1]]))
        a (sample (gamma 1 1))
        b (sample (gamma 1 1))
        d (dist a b)]
    ;; Data - samples from the distribution
    (observe d 1)
    (observe d 2)
    (observe d 4)
    (observe d 7)
    ;; Output --- predicted distribution
    (predict :d (type d))
    (predict :a a) (predict :b b)))
\end{lstlisting}

Internally, an Anglican query is represented by a computation in
\textit{continuation passing style} (CPS), and inference algorithms
exploit the CPS structure of the code to intercept probabilistic
operations in an algorithm-specific way. Among the available
inference algorithms there are Particle Cascade~\cite{PWD+14},
Lightweight Metropolis-Hastings~\cite{WSG11}, Iterative Conditional
Sequential Monte-Carlo (Particle Gibbs)~\cite{WVM14}, and others.
Inference on Anglican queries generates a lazy sequence of
samples, which can be processed asynchronously in Clojure
code for analysis, integration, and decision making.

@@

Anglican can be mostly implemented as a regular programming
language, except for the handling of these checkpoints.
Depending on the \textit{inference algorithm}, \texttt{sample}
and \texttt{observe} may result in implicit input/output
operations and control changes. For example, \texttt{observe} in
particle filtering inference algorithms is a non-deterministic
control statement at which a particle can be either replicated
or terminated. Similarly, in Metropolis-Hastings,
\texttt{sample} is both an input and a non-deterministic control
statement (with delayed effect), eventually affecting acceptance
or rejection of a sample.

@@

Compilation, invocation, and runtime support of Anglican queries
are discussed in detail in further sections.

@@

Providing an explicit syntax for differentiating between
Anglican and Clojure function calls would be cumbersome. Another
option would be to use meta-data to identify Anglican function
calls at runtime, however this would impact performance, and
a good runtime performance is critical for probabilistic
programs. The approach taken by Anglican is to maintain a list
of unqualified names of primitive functions, as well of
namespaces in which all functions are primitive, and recognize
primitive functions by name --- if a function name is not in the
list, the function is an Anglican function. Global dynamically
bound variables \texttt{*primitive-procedures*} and
\texttt{*primitive-namespace*} contain the initial lists of
names and namespaces, correspondingly. Of course, local bindings
can shade global primitive function names. For example,
\texttt{first} is a Clojure function inside the let block in the
following example:
\begin{lstlisting}[style=default]
(let [first (fn [[x & y]] x)]
  (first '[1 2 3]))
\end{lstlisting}
The Anglican compiler takes care of the shading by rebinding
\texttt{*primitive-procedures*} in every lexical scope
(\texttt{fn-cps}, \texttt{cps-of-let}). Macro
\texttt{shading-primitive-procedures} automates the shading.

>>>>

While probabilistic programming systems shorten the iteration cycle in
exploratory model design, they typically lack basic functionality needed for
data I/O, pre-processing, and analysis and visualization of inference results.
In this demonstration, we describe the implementation of
Anglican~(\cite{TMW15}), a probabilistic
programming language that tightly integrates with
Clojure~(\url{http://clojure.org/}), a general-purpose programming language
that runs on the Java Virtual Machine (JVM). Both languages share a common
syntax, and can be invoked from each other. This allows Anglican programs to
make use of a rich set of libraries written in both Clojure and Java.
Conversely Anglican allows intuitive and compact specification of models for
which inference may be performed as part of a larger Clojure project.

There are several ways to build a programming language
on top or besides another language.  The easiest to grasp is an
interpreter --- a program that reads a program, in its entirety
or line-by-line, and executes it by applying operational
semantics of a certain kind to the language. \textsc{Basic} is
famous for line-by-line interpreted implementations.

@@

An implementation of Anglican must therefore address three issues:
\begin{itemize}
    \item the Clojure syntax to introduce probabilistic Anglican
        code inside Clojure modules;
    \item source-to-source transformation of Anglican programs
        into Clojure, so that probabilistic execution becomes
        possible;
    \item algorithms which run Clojure code, obtained by
        transforming Anglican programs, according to the
        probabilistic operational semantics.
\end{itemize}

Anglican language syntax, compilation, invocation, and runtime
support of Anglican queries are discussed in detail in further
sections.

@@

An Anglican program, or \textit{query}, is compiled into a Clojure
function. When inference is performed with a provided algorithm, this
produces a sequence of output values, or \textit{predicts}. Anglican
shares a common syntax with Clojure; Clojure functions can be called
from Anglican code and vice versa. A simple program in Anglican can
look like the following code:
\begin{lstlisting}[style=default]
(defquery models data
  "chooses a distribution
   which describes the data"
  (let [;; Guessing a distribution
        dist (sample (categorical
                       [[uniform-continuous 0 1]
                        [normal 0.5 1] 
                        [gamma 1 1]]))
        a (sample (gamma 1 1))
        b (sample (gamma 1 1))
        d (dist a b)]

    ;; Observing samples from the distribution
    (loop [data data]
      (when (seq data)
        (let [[x & data] data]
          (observe d x))
        (recur data)))

    ;; Predicting the distribution
    (predict :d (type d))
    (predict :a a) (predict :b b)))
\end{lstlisting}

Internally, an Anglican query is represented by a computation in
\textit{continuation passing style} (CPS), and inference algorithms
exploit the CPS structure of the code to intercept probabilistic
operations in an algorithm-specific way\footnote{\cite{GS15}
also describe a CPS-based implementation of a probabilistic
programming language.}. Among the available
inference algorithms there are Particle Cascade~\cite{PWD+14},
Lightweight Metropolis-Hastings~\cite{WSG11}, Iterative Conditional
Sequential Monte-Carlo (Particle Gibbs)~\cite{WVM14}, and others.
Inference on Anglican queries generates a lazy sequence of
samples, which can be processed asynchronously in Clojure
code for analysis, integration, and decision making.

@@

Anglican can be mostly implemented as a regular programming
language, except for the handling of these checkpoints.
Depending on the \textit{inference algorithm}, \texttt{sample}
and \texttt{observe} may result in implicit input/output
operations and control changes. For example, \texttt{observe} in
particle filtering inference algorithms \cite{WVM14} a non-deterministic
control statement at which a particle can be either replicated
or terminated. Similarly, in Metropolis-Hastings \cite{WSG11},
\texttt{sample} is both an input and a non-deterministic control
statement (with delayed effect), eventually affecting acceptance
or rejection of a sample.

@@

@@

Providing an explicit syntax for differentiating between
Anglican and Clojure function calls would be cumbersome. Another
option would be to use meta-data to identify Anglican function
calls at runtime, however this would impact performance, and
a good runtime performance is critical for probabilistic
programs. The approach taken by Anglican is to maintain a list
of unqualified names of primitive functions, as well of
namespaces in which all functions are primitive, and recognize
primitive functions by name --- if a function name is not in the
list, the function is an Anglican function. Global dynamically
bound variables \texttt{*primitive-procedures*} and
\texttt{*primitive-namespace*} contain the initial lists of
names and namespaces, correspondingly. Of course, local bindings
can shade global primitive function names. For example,
\texttt{first} is an Anglican function inside the let block in the
following example:
\begin{lstlisting}[style=default]
(let [first (fn [[x & y]] x)]
  (first '[1 2 3]))
\end{lstlisting}
The Anglican compiler takes care of the shading by rebinding
\texttt{*primitive-procedures*} in every lexical scope
(\texttt{fn-cps}, \texttt{cps-of-let}). Macro
\texttt{shading-primitive-procedures} automates the shading.
